\DocumentMetadata{}
\documentclass[a4paper,12pt]{article}
\usepackage{xspace}
\usepackage{palatino}

\newcommand{\nm}{Diyrbal\nm}

\begin{document}

\section*{Preface}

\nm is a higher-order, high-level, type-safe, dynamically-typed, multi-paradigm, UNIX-native but portable scripting and programming language. It has been influenced, both in sytnax and semantics, and also in implementation, by the likes of Ruby, Perl, Scheme, Tcl and Python.

It's best to introduce \nm with a basic example. The factorial function should fit the occasion:

\begin{diyrbalcode}
	let fac <- fnrec
		begin 
			if !\$1 return 0; 
			else return fac(\$1 - 1) * fac(\$1 - 2) 
		end
\end{diyrbalcode}

\textbf{Do not} let the tacit\footnote{Meaning, the function definition does not specify formal parameters, and actual parameters -- i.e. arguments, are accessed by \texttt{\$1}} nature of above code turn you off the language! It also allows named formal parameters. We'll see that later.

% TODO introduce the language more

\section{Features of \nm}

\nm boasts several multi-paradigm features found in other languages, or discussed in academic literature. Though it is a fully imperative language, functional programming is not fallen by the wayside in \nm. It is a big selling point for it. Since I plan on writing compilers with \nm, I wish to implement features that make verification of the code easier. One such feature is the \textit{Algebraic effect}.

Algebraic effects are an alternative to monads for handling side-effects and stateful modifications. Within an Algebraic effect, one specifies an \textit{effect}, and then bundles that effect with state-altering constructs, and lets the \textit{handler} he specifies to take care of the effect.

Another feature sported by \nm is built-in lexer/parser in form of \textit{grammars}. If your Perl 6 Vietnam flashbacks are going off right now, please keep in mind that grammars have always been an strong suite of the language, and \nm wishesh to follow in its footsteps by providing the user with a ready-made intrinsic lexer/parser.

The interpreter also provides functional-style pattern-matching constructs. It might not be as modern as Algebraic effects, since pattern-matching has been around since the 70s, introduced by ML. But a modern language cannot survive without pattern matching!

To accompany pattern-matching, \nm provides sum, product, and sum-of-product types, collectively known as \textit{Algebraic data types}. In other words, records and variants, and variants containing records. Because of this, \nm does not feel the need to provide imperative-style enumeration types. In an scripting language, it's more likely that the programmer would use strings to enumerate anyways. \texttt{Option} and \texttt{Result} types are built-in into the language, this might sound inexusable, but the syntax for Option and Result differ from that of sum-of-product types used by other languages. So I stand by this choice!

The language also supports \textit{polymorphism}. Not the OOP-kind of polymorphism, rather, the system infuenced by System F\citework{tapl}. However, it is my belief that a \textit{faint} amount of OOP is necessary in a language that is targeted for both scripting, as well as scripting. So it supports \textit{roles} and \textit{protocols}. The roles are similar to typeclasses\citework{haskellstd} and protocols are similar to interfaces\citework{gostd}. The former is provided for functional programming, the latter, for imperative.

\nm provides \textit{Green threads} for all its prallelism, and concurrency needs. Green threads are virtual threads, scheduled by the runtime, rather than the OS. \textit{However}, the backend decides to elavate a Green thread to an actual kernel-managed thread (or process) whenever it desires. If the user increases the maximum number of concurrent threads, which defaults to one, the runtime will comply and use kernel-managed threads to spread the scheduler across.

Concurrency in \nm is spread around \textit{generators}, \textit{coroutines}, and \textit{coprocesses}. Generators are similar to Python generators, coroutines are asyncronous functions that return a promise, and coprocesses launch an entirely new instance of the interpreter, that uses IP methods to communicate with the current one.

The language also supports chaining of expressions via piping, a la OCaml (\texttt{e\textsubscript{1} |> e\textsubscript{2}}), and composition of function, a la Haskell (\texttt{f\textsubscript{1} . f\textsubscript{2}}).

\end{document}
